Gradual Soundness: Lessons from Static Python

### 2022-01-04 pj-7.1/paper.tex
pj: sec 3, table describing run-time checks
purpose is to show that checks are simple and complete

\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\Gamma}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{#1(#2)\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}

@@

\newcommand{\mfapply}[2]{#1\,(#2)}
\newcommand{\mffont}[1]{\mathit{#1}}
\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
\newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}

@@

@@

    \inferrule*{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
    }

@@

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_1
    }{
      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptdyn
    }

@@

\begin{figure}[t]
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

  \caption{TBA Surface Types to Evaluation Types}
  \label{f:surface-to-eval-types}
\end{figure}

@@

>>>>

\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\Gamma}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{#1(#2)\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}

@@

\newcommand{\mfapply}[2]{#1\,(#2)}
\newcommand{\mffont}[1]{\mathit{#1}}
\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
\newcommand{\mfcast}[2]{\mfapply{\mffont{Cast}}{#1, #2}}
\newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}

@@

%% 2022-01-04 argument = nom soundness -> sp model sound -> sp sound via conformance equivalence
%% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)

%% TODO 2022-01-04 (somewhere) explain that a verified compiler is an entirely separate project,
%%  need bytecode model, full transformations, memory model / machine etc.
%%  hence we don't model optimizations, little benefit
%%  opt. don't tell us about the surface language ... the particular SP opts are changing representations but not results

@@

    \inferrule*{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
    }

@@

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptdyn
    }

@@

\begin{figure}[t]
  \begin{langarray}
    \spvalue & \defeq &
      \mbox{heap-allocated Python values}
    \\
    \spteval & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptobject \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
    \\ & &
      \sptrawpydict \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptoptional{\sptype}
  \end{langarray}

  \bigskip
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

  \caption{TBA Type Metafunctions}
  \label{f:surface-to-eval-types}
\end{figure}

\begin{table}
  \caption{TBD Run-time checks}
  \label{t:cast}

  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any non-primitive value} \\  %% except primitives
    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (\code{0} and \code{1})} \\
    \sptfloat & \mbox{Accepts floats}
  \end{array}\)
\end{table}

@@

%% 2022-01-04 TODO ask C&D how many teams have adopted SP ... get sense of adoption
